Staged Abstract Interpreters

### 2018-11-11 paper/22_bg_abs.tex
work on paper

Based on the generic interface we presented, now we will describe how to instantiate an abstract interpreter.
To keep the simplicity, we intened to use simply abstract domain, and just establish a context-insensitive, 
path-insensitive, and flow-insensitive analysis in this section -- it is coarse but enough to lay a foundation for 
the staged abstract interpreter we will present later.
In Section \note{X}, we will see how to regain 
context-/path-/flow-sensitivity for control-flow analysis; in Section \note{Y}, we will see how to 
instantiate an interval abstract domain for numerical analysis.

Following the idea of AAM, store widened

\textbf{Abstract Instantiation.}
We use powerset as our abstract domain, the concrete closures directly becomes elements in the powerset, numbers
are abstracted to a single abstract number element \texttt{NumV}

@@

\textbf{Stage Polymorphic Lattices.}
We use type class in Scala, also works as stage polymorphic.
\begin{lstlisting}
trait Lattice[E, R[_]] {
  val bot: R[E];  val top: R[E]
  def $\sqsubseteq$(l1: R[E], l2: R[E]): R[Boolean]
  def $\sqcup$(l1: R[E], l2: R[E]): R[E]
  def $\sqcap$(l1: R[E], l2: R[E]): R[E]
}
\end{lstlisting}

For example, the abstract domain we use here, power set can be implemented as:

@@

tuples, maps can be lifted to lattice element-wise or point-wise, the code for them are elided.

\todo{conditional, should use different store}

\begin{lstlisting}
object AbsInterp extends Abstract {
  type R[+T] = T
  val $\rho$0 = Map[Ident, Addr]();  val $\sigma$0 = Map[Addr, Value]()
  def get($\rho$: Env, x: Ident): Addr = $\rho$(x)
  def put($\rho$: Env, x: Ident, a: Addr): Env = $\rho$ + (x -> a)
  def get($\sigma$: Store, a: Addr): Value = 
    $\sigma$.getOrElse(a, Lattice[Value].bot)
  def put($\sigma$: Store, a: Addr, v: Value): Store =
    $\sigma$ + (a -> (v $\sqcup$ get($\sigma$, a)))
  def alloc($\sigma$: Store, x: Ident): Addr = Addr(x)
  def close(ev: EvalFun)($\lambda$: Lam, $\rho$: Env): Value = Set(CloV($\lambda$, $\rho$))
  def num(i: Lit): Value = Set(NumV())
  def apply_closure(ev: EvalFun)(f: Value, arg: Value, $\sigma$: Store): Ans = {
    var $\sigma$0 = $\sigma$
    val vs = for (CloV(Lam(x, e), $\rho$) <- f) yield {
      val $\alpha$ = alloc($\sigma$0, x)
      val $\rho$_* = put($\rho$, x, $\alpha$)
      val $\sigma$_* = put($\sigma$0, $\alpha$, arg)
      val (v, v$\sigma$) = ev(e, $\rho$_*, $\sigma$_*)
      $\sigma$0 = v$\sigma$; v
    }
    (vs.reduce(Lattice[Value].$\sqcup$(_,_)), $\sigma$0)
  }
  def branch0(cnd: Value, thn: => Ans, els: => Ans): Ans = {
    thn $\sqcup$ els //TODO!
  }
  def prim_eval(op: Symbol, v1: Value, v2: Value): Value = Set(NumV())
} // to be continued
\end{lstlisting}

>>>>

Based on the generic interface we presented in Section~\ref{bg_lang}, now we describe the instantiation of an
abstract interpreter.
To keep the simplicity, we intend to use simply abstract domains, and just establish a 
context-/path-/flow-insensitive, store-widened analysis in this section --- indeed, it is coarse 
but enough to setup a foundation for the staged abstract interpreter we will present later.
However, in Section \note{X}, we will see how to regain context-/path-/flow-sensitivity 
for pushdown control-flow analysis; in Section \note{Y}, we will see how to instantiate 
an interval abstract domain for numerical analysis.

\textbf{Abstract Instantiation.}
Our instantiation roughly follows the Abstracting Abstract Machines methodology 
\cite{DBLP:conf/icfp/HornM10, DBLP:journals/jfp/HornM12} that transforms the concrete semantics 
to abstract semantics. The store now maps addresses to sets of abstract values, meaning that an address points
to all possible values may occur at runtime, where the abstract value is either a closure or 
a single abstract number element \texttt{NumV}. The address space is also
constrained to be finite, in the monovariant setting, we simply use variable names for it.

@@

\textbf{Stage Polymorphic Lattices.}
To effectively reuse the code between unstaged and staged part, we also make the lattice structure to
be stage polymorphic. A lattice type class parameterizes over an element type \texttt{E} and defines
five operations for \texttt{R[E]}, such as meet, join, and ordering relation; again, \texttt{R} 
is the higher-kinded type indicating the binding-time.
\begin{lstlisting}
trait Lattice[E, R[+_]] {
  val bot: R[E];  val top: R[E]
  def $\sqsubseteq$(l1: R[E], l2: R[E]): R[Boolean]
  def $\sqcup$(l1: R[E], l2: R[E]): R[E]
  def $\sqcap$(l1: R[E], l2: R[E]): R[E]
}
\end{lstlisting}

For example, here we use powerset as the abstract domain, and the unstaged lattice 
of powerset can be easily implemented as follows:

@@

The ordering relation is to ask whether one set is a subset of the other, 
join is to union the two set, and meet is to intersect the two sets.
Accordingly, other structures used in the rest of the paper such as tuples 
(products) and maps can be lifted to lattices element-wise or point-wise.
The code for them are elided.

\textbf{Operations Implementation.}

The @get@ and @put@ operation on abstract stores slightly changed according to
the abstract semantics: @get@ uses bottom element in the @Value@ lattice if the queried
address not exists; @put@ performs the update by joining with the existing values.
@alloc@ simply invokes the @Addr@ constructor.
\texttt{close} and \texttt{num} lift syntactic literals to a singleton set that
contains one \texttt{CloV} or \texttt{NumV} object.

\todo{conditional, should use different store}
\begin{lstlisting}
object AbsInterp extends Abstract {
  type R[+T] = T
  val $\rho$0 = Map[Ident, Addr]();  val $\sigma$0 = Map[Addr, Value]()
  def get($\rho$: Env, x: Ident): Addr = $\rho$(x)
  def put($\rho$: Env, x: Ident, a: Addr): Env = $\rho$ + (x -> a)
  def get($\sigma$: Store, a: Addr): Value = 
    $\sigma$.getOrElse(a, Lattice[Value].bot)
  def put($\sigma$: Store, a: Addr, v: Value): Store =
    $\sigma$ + (a -> (v $\sqcup$ get($\sigma$, a)))
  def alloc($\sigma$: Store, x: Ident): Addr = Addr(x)
  def close(ev: EvalFun)($\lambda$: Lam, $\rho$: Env): Value = Set(CloV($\lambda$, $\rho$))
  def num(i: Lit): Value = Set(NumV())
  def apply_closure(ev: EvalFun)
    (f: Value, arg: Value, $\sigma$: Store): Ans = { var $\sigma$_* = $\sigma$
      val vs = for (CloV(Lam(x, e), $\rho$) <- f) yield {
        val $\alpha$ = alloc($\sigma$_*, x)
        val (v, v$\sigma$) = ev(e, put($\rho$, x, $\alpha$), put($\sigma$_*, $\alpha$, arg))
        $\sigma$_* = v$\sigma$; v
      }
    (vs.reduce(Lattice[Value].$\sqcup$), $\sigma$_*)
  }
  def branch0(cnd: Value, thn: =>Ans, els: =>Ans): Ans = thn $\sqcup$ els //TODO!
  def prim_eval(op: Symbol, v1: Value, v2: Value): Value = Set(NumV())
} // to be continued
\end{lstlisting}
